Ore Polynomials in Sage

### 2013-06-15 paper/main.tex
paper: more stuff added to section 4; gaps closed.

\documentclass{amsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}
\usepackage{sagetex}
\usepackage{tikz}

@@

Algorithms for Ore algebras provide a systematic answer to this
question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
century~\cite{ore33} with the objective of providing a unified theory for
various kinds of linear operators, they have been used for many years in
computer algebra systems, for example in the Maple packages
OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, in the
Mathematica packages by Mallinger~\cite{mallinger96} and
Koutschan~\cite{koutschan10c,koutschan09}.

@@

At the time of writing, the package we describe here is still under construction
and has not yet been incorporated into the official Sage distribution. Readers
who want to try it out are invited to download the current latest version from 
[URL] and are encouraged to send us bug reports and other comments. We hope 
that the community will find the code useful.

@@

\begin{sageexample}
  sage: #load("ore_algebra.py", "guessing.py", "ore_operator.py");  #### TO BE FIXED ####
  sage: from ore_algebra import *
  sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
  sage: A
  sage: A.random_element()
  sage: Dx*x
  sage: B.<Sx> = OreAlgebra(R)
  sage: B
  sage: Sx*x
\end{sageexample}

More details on the construction of Ore algebras are given in the following section.
Also Ore algebras with several generators are already supported. However, so far we offer
no functionality beyond addition and multiplication for these. More functionality 
is available for algebras with one generator. We describe some of it in 
Section~\ref{sec:3}. Even more functionality has been implemented for special 
algebras like the ones created above, which are most frequently used in practice. 
Section~\ref{sec:4} describes some of these features.

/// disclaimer: we don't explain algorithms here, but only the software ///

@@

\begin{center}
  \begin{tabular}{|l|p{.55\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|associated_commutative_algebra()| & returns a polynomial ring with the
       same base ring as $A$ and whose generators are named like the generators
       of~$A$\\
    \verb|base_ring()| & returns the base ring of $A$\\
    \verb|delta(i)| & returns a callable object representing the delta map
       associated to the $i$th generator (default: $i=0$) \\
    \verb|gen(i)| & returns the $i$th generator (default: $i=0$)\\
    \verb|sigma(i)| & returns a callable object representing the sigma map
       associated to the $i$th generator (default: $i=0$) \\
    \verb|var(i)| & returns the name of the $i$th generator (default: $i=0$)\\\hline
  \end{tabular}
\end{center}

@@

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
  sage: Ac = A.associated_commutative_algebra(); Ac
  sage: A(Ac.random_element())
  sage: A([5*x,7*x-3,3*x+1])
  sage: A("(5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)")
\end{sageexample}

@@

In accordance with the default Sage coercion model, such conversions take place automatically (if possible) 
when operators from different algebras are added or multiplied. Note that the result of such an operation
need not belong to either of the parents of the operands but may instead have a suitable ``common extension'' 
as parent.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx'); 
  sage: B = OreAlgebra(QQ['x'].fraction_field(), 'Dx')
  sage: L = A.random_element() + B.random_element()
  sage: L.parent()
\end{sageexample}

@@

\subsection{gcrd and lclm}

A univariate Ore algebra over a field is a left Euclidean domain, which means that it is possible to perform 
left division with remainder. Building upon this, the greatest common right divisor (GCRD) and the least common 
left multiple (LCLM) of two Ore polynomials can be computed. The package provides a number of methods to carry 
out these tasks

\begin{center}
  \begin{tabular}{|l|p{.55\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|quo_rem(B)| & returns the left quotient and the left remainder of $A$ and $B$.\\
    \verb|gcrd(B)| & returns the greatest common right divisor of $A$ and $B$.\\
    \verb|xgcrd(B)| & returns the greatest common right divisor of $A$ and $B$ and the according B\'ezout coefficients.\\
    \verb|lclm(B)| & returns the least common left multiple of $A$ and $B$.\\\hline
  \end{tabular}
\end{center}

All these methods are also available for Ore operators living in univariate Ore algebras over a base ring $R$ which 
does not necessarily have to be a field. The operators are then implicitly assumed to live in the respective Ore 
algebra over the quotient field $K$ of $R$. The output will be the GCRD (LCLM, quotient, remainder) in the Ore 
algebra over $K$ but not over $R$, in which these objects might not exist or might not be computable.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['n'], 'Sn');                                           
  sage: G,L1,L2 = A.random_element(2),A.random_element(7),A.random_element(5)    
  sage: while not L1.gcrd(L2)==1: L2 = A.random_element(5)                       
  sage: L1,L2 = L1*G,L2*G                                                        
  sage: (L3,S,T) = L1.xgcrd(L2)                             
  sage: (S*L1+T*L2).normalize() == L3.normalize() == G.normalize()
  sage: LCLM = L1.lclm(L2)
  sage: LCLM.quo_rem(L1)[1] == LCLM.quo_rem(L2)[1] == A.zero()
\end{sageexample}

There are a number of options to adjust the functionality of the methods. For example, the efficiency of computing the GCRD 
depends on the size of the coefficients of intermediate results and there are different strategies to control this in so called 
polynomial remainder sequences (PRS). The default is the improved PRS described in \cite{jaroschek13b}. which will usually be 
the fastest choice.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['n'], 'Sn');                                           
  sage: L1,L2 = A.random_element(3),A.random_element(2)    
  sage: #L1.gcrd(L2, prs=__primitivePRS__)
  sage: #L1.gcrd(L2, prs=__classicPRS__)
\end{sageexample}

\subsection{closure properties?}

-- table mapping function names to arithmetic

-- example for how to prove some identity with the methods provided.

-- merge with previous section?

\subsection{conversions?}

-- re2de and all its relatives

\subsection{solutions?}

-- polynomial, rational, generalized series solutions

-- available for differential and shift algebras

\subsection{action?}

-- should be mentioned somewhere, although there is not much to say about it.

\subsection{Differential Operators}

Useful methods available for differential operators:

\begin{center}
  \begin{tabular}{|l|p{.55\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|to_S()| & converts to a recurrence operator for the Taylor series solutions at the origin \\
    \verb|to_F()| & converts to a difference operator for the Taylor series solutions at the origin \\
    \verb|to_T()| & rewrites in terms of the Euler derivative \\
    \verb|annihilator_of_integral()| & converts an annihilator for $f(x)$ to one for $\int f(x) dx$\\
    \verb|annihilator_of_composition()| & converts an annihilator for $f(x)$ to one for $f(a(x))$ where $a(x)$ is algebraic over the base ring\\
    \verb|power_series_solutions()| & computes power series solutions of this operator\\
    \verb|generalized_series_solutions()| & computes generalized series solutions of this operator \\ \hline
  \end{tabular}
\end{center}

@@

We compute an annihilator for the error function $\frac{2}{\sqrt{\pi}} \int_0^x e^{-t^2} dt$,
starting from the differential equation for $e^x$, and produce the
recurrence for the Taylor series coefficients at the origin:

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R, 'Dx')                            
  sage: (Dx - 1).annihilator_of_composition(-x**2)
  Dx + 2*x
  sage: (Dx + 2*x).annihilator_of_integral()                                     
  Dx^2 + 2*x*Dx
  sage: (Dx^2 + 2*x*Dx).to_S(OreAlgebra(ZZ['n'], 'Sn'))                          
  (n^2 + 3*n + 2)*Sn^2 + 2*n
  sage: (Dx^2 + 2*x*Dx).power_series_solutions(10)                               
  [x - 1/3*x^3 + 1/10*x^5 - 1/42*x^7 + 1/216*x^9 + O(x^10), 1 + O(x^10)]
\end{sageexample}

\subsection{Computation of Sequence Terms}

(... a few introductory words ... available for shift and q-shift algebras ...)

Useful methods available for recurrence operators:

\begin{center}
  \begin{tabular}{|l|p{.55\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|to_D()| & converts annihilator for the coefficients in a power series to a
        differential operator for the sum \\
    \verb|to_F()| & converts shift operator to a difference operator\\
    \verb|to_T()| & converts to a differential operator in terms of the Euler derivative \\
    \verb|annihilator_of_sum()| & converts an annihilator for $f(n)$ to one for $\sum_{k=0}^n f(k)$\\
    \verb|annihilator_of_composition()| & converts an annihilator for $f(n)$ to one for $f(\lfloor un+v \rfloor)$ where $u, v \in \mathbb{Q}$ \\
    \verb|annihilator_of_interlacing()| & interlaces two or more sequences\\
    \verb|generalized_series_solutions()| & computes asymptotic expansions of sequences annihilated by the operator\\ \hline
  \end{tabular}
\end{center}

@@

We compute an annihilator for the sequence
$c(n) = \sum_{k=0}^n 1 / k!$:

\begin{sageexample}
  sage: R.<n> = ZZ[]; RS.<Sn> = OreAlgebra(R)
  sage: inverse_factorials = (n + 1) * Sn - 1
  sage: partial_sums = inverse_factorials.annihilator_of_sum()
  sage: partial_sums
\end{sageexample}

@@

... add a figure with timings? ...

\subsection{The $q$-Case}

J and Q

\section{Guessing}\label{sec:5}

[[CITE RUBEY.]]

Guessing is one of the most popular features of software packages like gfun.  It
is a technique for detecting patterns in sequences of numbers, which is often
used in combinatorics. The basic idea is simple. If we are given a finite array
of numbers, though of as the first terms of an infinite sequence, we may want to
know whether this sequence satisfies a recurrence. The algorithm behind a
guessing engine searches for small equations matching the given data.
Generically, no such equations exist, so if some are found, it is fair to
``guess'' that they are in fact valid equations for the whole infinite sequence.
=======
\subsection{Guessing}

Guessing is, in some sense, the reverse operation of \verb|to_list|. It is one
of the most popular features of packages like gfun, and there are even some
special purpose packages dedicated to this technique~\cite{kauers09a,hebisch11}.
The basic idea is simple. If we are given a finite array of numbers, thought of
as the first terms of an infinite sequence, we may want to know whether this
sequence satisfies a recurrence. The algorithm behind a guessing engine searches
for small equations matching the given data.  Generically, no such equations
exist, so if some are found, it is fair to ``guess'' that they are in fact valid
equations for the whole infinite sequence.
>>>>>>> manuel/master

@@

The method tests a sequence of points $(r_1,d_1)$, $(r_2,d_2)$, \dots\ right
below this second parabola. Success at a point $(r_i,d_i)$ means that some
evidence for an operator of order $\leq r_i$ and degree $\leq d_i$ has been found. 
This operator however is not explicitly computed. Instead, the method uses the
partial information found about this operator to calculate an operator which with 
high probability is the minimal order operator satisfied by the sequence in question. 
This operator is usually more interesting than the one at $(r_i,d_i)$, and its 
computation is usually more efficient.

@@

Finally, there is an option \verb|infolevel| which, when set to a positive integer, causes the method
to print some output about the progress of the computation. Higher values lead to more detailed output.

\section{Concluding Remarks}

...

>>>>

\documentclass[11pt]{amsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}
\usepackage{sagetex}
\usepackage{tikz}

@@

Algorithms for Ore algebras provide a systematic answer to this
question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
century~\cite{ore33} with the objective of providing a unified theory for
various kinds of linear operators, they have been used for many years in
computer algebra systems, for example in the Maple packages
OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, or in
the Mathematica packages by Mallinger~\cite{mallinger96} and
Koutschan~\cite{koutschan10c,koutschan09}.

@@

At the time of writing, the package we describe here is still under construction
and has not yet been incorporated into the official Sage distribution. Readers
who want to try it out are invited to download the current latest version from 
\begin{center}
  URL
\end{center}
and are encouraged to send us bug reports and other comments. We hope 
that the community will find the code useful.

@@

\begin{sageexample}
  sage: from ore_algebra import *
  sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
  sage: A
  sage: A.random_element()
  sage: Dx*x
  sage: B.<Sx> = OreAlgebra(R)
  sage: B
  sage: Sx*x
\end{sageexample}

More details on the construction of Ore algebras are given in the following
section. The construction and manipulation of elements of Ore algebras is
discussed in Section~\ref{sec:3}.

The package also supports Ore algebras with several generators already. However,
so far we offer no functionality beyond addition and multiplication for
these. Much more functionality is available for algebras with one generator. We
describe some of it in Section~\ref{sec:4}.

@@

\begin{center}
  \begin{tabular}{|l|p{.5\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|associated_commutative_algebra()| & returns a polynomial ring with the
       same base ring as $A$ and whose generators are named like the generators
       of~$A$\\
    \verb|base_ring()| & returns the base ring of $A$\\
    \verb|delta(i)| & returns a callable object representing the delta map
       associated to the $i$th generator (default: $i=0$) \\
    \verb|gen(i)| & returns the $i$th generator (default: $i=0$)\\
    \verb|sigma(i)| & returns a callable object representing the sigma map
       associated to the $i$th generator (default: $i=0$) \\
    \verb|var(i)| & returns the name of the $i$th generator (default: $i=0$)\\\hline
  \end{tabular}
\end{center}

\smallskip\goodbreak

@@

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
  sage: Ac = A.associated_commutative_algebra()
  sage: Ac
  sage: A(Ac.random_element())
  sage: A([5*x,7*x-3,3*x+1])
  sage: A("(5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)")
\end{sageexample}

@@

In accordance with the Sage coercion model, such conversions take place automatically (if possible) 
when operators from different algebras are added or multiplied. Note that the result of such an operation
need not belong to either of the parents of the operands but may instead have a suitable ``common extension'' 
as parent.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx')
  sage: B = OreAlgebra(QQ['x'].fraction_field(), 'Dx')
  sage: L = A.random_element() + B.random_element()
  sage: L.parent()
\end{sageexample}

@@

Besides basic arithmetic for Ore operators, the package provides a wide range of
methods to create, manipulate and solve several different kinds of operators.
Some of these methods are accessible in any Ore algebra while others are tied
specifically to e.g.\ recurrence operators or differential operators.

In this section, we give an overview of some of the functionality provided by
the package. Because of space limitation, only some of the available methods
can be discussed here. For further information, we refer to the documentation.

\subsection{Methods for General Algebras}

A univariate Ore algebra over a field is a left Euclidean domain, which means
that it is possible to perform left division with remainder. Building upon this,
the greatest common right divisor (GCRD) and the least common left multiple
(LCLM) of two Ore polynomials can be computed. The package provides a number of
methods to carry out these tasks.

\begin{center}
  \begin{tabular}{|l|p{.65\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|A.quo_rem(B)| & returns the left quotient and the left remainder of $A$ and $B$.\\
    \verb|A.gcrd(B)| & returns the greatest common right divisor of $A$ and $B$.\\
    \verb|A.xgcrd(B)| & returns the greatest common right divisor of $A$ and $B$ and the according B\'ezout coefficients.\\
    \verb|A.lclm(B)| & returns the least common left multiple of $A$ and $B$.\\
    \verb|A.xlclm(B)| & returns the least common left multiple $L$ of $A$ and $B$ and the left quotients of $L$ and $A$ and of $L$ and $B$.\\
    \verb|A.resultant(B)| & returns the resultant of $A$ and $B$ (see \cite{li96} for further information).\\
    \hline
  \end{tabular}
\end{center}

All these methods are also available for Ore operators living in univariate Ore
algebras over a base ring $R$ which does not necessarily have to be a
field. The operators are then implicitly assumed to live in the respective Ore
algebra over the quotient field $K$ of~$R$. The output will be the
GCRD (LCLM, quotient, remainder) in the Ore algebra over $K$ but not over
$R$, in which these objects might not exist or might not be computable.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['n'], 'Sn')
  sage: G = A.random_element(2)
  sage: L1, L2 = A.random_element(7), A.random_element(5)
  sage: while L1.gcrd(L2) != 1: L2 = A.random_element(5)                       
  sage: L1, L2 = L1*G, L2*G                                                        
  sage: L1.gcrd(L2) == G.normalize()
  sage: L3, S, T = L1.xgcrd(L2)                             
  sage: S*L1 + T*L2 == L3
  sage: LCLM = L1.lclm(L2)
  sage: LCLM % L1 == LCLM % L2 == 0
  sage: LCLM.order() == L1.order() + L2.order() - G.order()
\end{sageexample}

The GCRD is only unique up to multiplication (from the left) by an element from
the base ring. The method \verb|normalize| called in line~6 of the listing above
multiplies a given operator from the left by some element from the base ring
such as to produce a canonical representative of the set of all the operators
that can be obtained from the given one by left multiplication of a base ring
element. This facilitates the comparison of output.

There are a number of options to adjust the functionality of the methods. For
example, the efficiency of computing the GCRD depends on the size of the
coefficients of intermediate results and there are different strategies to
control this in so called polynomial remainder sequences (PRS). The default is
the improved PRS described in~\cite{jaroschek13b}. which will usually be the
fastest choice.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['n'], 'Sn')
  sage: L1, L2 = A.random_element(3), A.random_element(2)
  sage: algs = ["improved", "classic", "monic", "subresultant"]
  sage: [L1.gcrd(L2, prs=a) for a in algs]
\end{sageexample}

If $L_1,L_2$ are operators, then the solutions of their GCRD are precisely the
common solutions of $L_1$ and~$L_2$. The LCLM, on the other hand, is the minimal
order operator whose solution space contains all the solutions of $L_1$ as well
as all the solutions of~$L_1$. Because of this property, GCRD and LCLM are
useful tools for constructing operators with prescribed solutions. For example,
here is how to construct a differential operator which has the solutions $x^5$
and $\exp(x)$ starting from the obvious operators annihilating $x^5$
and~$\exp(x)$, respectively.

\begin{sageexample}
  sage: R.<x> = ZZ[]; A.<Dx> = OreAlgebra(R)
  sage: L = (Dx - 1).lclm(x*Dx - 5)
  sage: L
  sage: L(x^5)
  sage: L(exp(x)).full_simplify() 
\end{sageexample}

Observe how in the last two lines we apply the operator~$L$ to other
objects. Such applications are not defined for every algebra and in general have
to be specified by the user through an optional argument:

\begin{sageexample}
  sage: A.<Qqn> = OreAlgebra(ZZ['q']['qn'])
  sage: var('q', 'n', 'x')
  sage: (Qqn^2+Qqn+1)(q^n, action=lambda expr: expr.substitute(n=n+1))
  sage: (Qqn^2+Qqn+1)(x, action=lambda expr: expr.substitute(x=q*x))
\end{sageexample}

Thanks to the LCLM operation discussed above, we have the property that when $f$
and $g$ are two objects which are annihilated by some operators $L_1,L_2$
belonging to some Ore algebra~$A$ then this algebra contains also an operator
which annihilates their sum~$f+g$. Thus the class of solutions of operators of
$A$ is closed under addition. It turns out that similar closure properties hold
for other operations. The following table lists some of the corresponding
methods. Methods for some further closure properties will appear further below.

\begin{center}
\begin{tabular}{|l|p{.55\hsize}|}
\hline
Method name & short description \\\hline
\verb|lclm()| & computes an annihilating operator for $f+g$ from annihilating operators for $f$ and $g$ \\
\verb|symmetric_product()| & computes an annihilating operator for $fg$ from annihilating operators for $f$ and $g$ \\
\verb|symmetric_power()| & computes an annihilating operator for $f^n$ from an annihilating operator for $f$ and a given positive integer~$n$\\ 
\verb|annihilator_of_associate()| & computes an annihilating operator for $M(f)$ from an annihilating operator for $f$ and a given operator~$M$\\ 
\verb|annihilator_of_polynomial()| & computes an annihilating operator for the object $p(f,\partial f,\partial^2 f,\dots)$ from an annihilating operator for $f$ and a given multivariate polynomial~$p$. \\
\hline
\end{tabular}
\end{center}

As an example application, let us prove Cassini's identity for Fibonacci
numbers: 
\[
  F_{n+1}^2 - F_nF_{n+2}=(-1)^n.
\]
The idea is to derive, using commands
from the table above, a recurrence satisfied by the left hand side and then show
that this recurrence is also valid for the right hand side.

\begin{sageexample}
  sage: A.<Sn> = OreAlgebra(ZZ['n'])
  sage: fib = Sn^2 - Sn - 1
  sage: R.<x0,x1,x2> = ZZ['n']['x0','x1','x2']
  sage: fib.annihilator_of_polynomial(x1^2 - x0*x2) 
\end{sageexample}

As this operator obviously annihilates $(-1)^n$, the proof is complete after
checking that the identity holds for $n=0$. Another way of carrying out the
same computation using the other commands would be as follows.

\begin{sageexample}
  sage: A.<Sn> = OreAlgebra(ZZ['n'])
  sage: fib = Sn^2 - Sn - 1
  sage: L1 = fib.annihilator_of_associate(Sn).symmetric_power(2)
  sage: L2 = fib.annihilator_of_associate(Sn^2).symmetric_product(fib)
  sage: L1.lclm(L2)
\end{sageexample}

Observe that the resulting operator again annihilates~$(-1)^n$, but its order is
higher than the operator obtained before, so we need to check more initial
values to complete the proof. For larger computations, the command
\verb|annihilator_of_polynomial| would also consume less computation time than
the step-by-step approach.

\subsection{Methods for Special Algebras}

For the elements of some of the most important algebras, additional methods are
available. The following table lists some of the additional methods available
for differential operators, i.e., elements of an Ore algebra of the
form~$R[x]\<Dx>$ or $K(x)\<Dx>$.

\begin{center}
  \begin{tabular}{|l|p{.55\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|to_S()| & converts to a recurrence operator for the Taylor series solutions at the origin \\
    \verb|to_F()| & converts to a difference operator for the Taylor series solutions at the origin \\
    \verb|to_T()| & rewrites in terms of the Euler derivative \\
    \verb|annihilator_of_integral()| & converts an annihilator for $f(x)$ to one for $\int f(x) dx$\\
    \verb|annihilator_of_composition()| & converts an annihilator for $f(x)$ to one for $f(a(x))$ where $a(x)$ is algebraic over the base ring\\
    \verb|desingularize()| & computes a left multiple of this operator with polynomial coefficients and 
           lowest possible leading coefficient degree\\
    \verb|associate_solutions(p)| & applied to an operator~$P$, this computes, if possible, an operator $M$
    and a rational function $m$ such that $DM=p+mP$ (see \cite{abramov99} for further information)\\
    \verb|polynomial_solutions()| & computes the polynomial solutions of this operator\\
    \verb|rational_solutions()| & computes the rational function solutions of this operator\\
    \verb|power_series_solutions()| & computes power series solutions of this operator\\
    \verb|generalized_series_solutions()| & computes generalized series solutions of this operator \\ \hline
  \end{tabular}
\end{center}

@@

As an example application, we compute an annihilator for the error function
$\frac{2}{\sqrt{\pi}} \int_0^x \exp(-t^2) dt$, starting from the differential
equation for $\exp(x)$, and produce the recurrence for the Taylor series
coefficients at the origin. Finally, we compute the series solutions of the
differential operator at infinity.

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R, 'Dx')                            
  sage: (Dx - 1).annihilator_of_composition(-x^2)
  sage: L = (Dx + 2*x).annihilator_of_integral()                                     
  sage: L
  sage: L.to_S(OreAlgebra(ZZ['n'], 'Sn'))                          
  sage: L.power_series_solutions(10)                               
  [x - 1/3*x^3 + 1/10*x^5 - 1/42*x^7 + 1/216*x^9 + O(x^10), 1 + O(x^10)]
  sage: L.annihilator_of_composition(1/x).generalized_series_solutions()
\end{sageexample}

The last output implies that the operator annihilating $\int_0^x \exp(-t^2) dt$
admits also a solution which behaves for $x\to\infty$ like $\frac1x\exp(-x^2)$.

The next example illustrates the methods for finding rational and polynomial
solutions of an operator~$L$.  These methods accept as an optional parameter an
inhomogeneous part consisting of a list (or tuple) of base ring elements,
$(f_1,\dots,f_r)$. They return as output a list of tuples $(g,c_1,\dots,c_r)$
with $L(g)=c_1f_1+\cdots+c_rf_r$ where $g$ is a polynomial or rational function
and $c_1,\dots,c_r$ are constants, i.e., elements of the base ring's base
ring. The tuples form a basis of the solution space.

In the example session below, we start from two polynomials $p,q$, then compute an 
operator $L$ having $p$ and $q$ as solutions, and then recover $p$ and $q$ from~$L$.
Note that for consistency also the solutions of homogeneous equations are returned as 
tuples. At the end we give an example for solving an inhomogeneous equation.

\begin{sageexample}
  sage: R.<x> = ZZ[]
  sage: p = x^2 + 3*x + 8; q = x^3 - 7*x + 5
  sage: A.<Dx> = OreAlgebra(R)
  sage: L = (p*Dx - p.derivative()).lclm(q*Dx - q.derivative())
  sage: L
  sage: L.polynomial_solutions()
  sage: M = (2*x+3)*Dx^2 + (4*x+5)*Dx + (6*x+7)
  sage: sol = M.polynomial_solutions([1,x,x^2,x^3])
  sage: sol
  sage: map(lambda s: M(s[0]) == s[1]+s[2]*x+s[3]*x^2+s[4]*x^3, sol)
\end{sageexample}

The functions \verb|polynomial_solutions| and \verb|rational_solutions| are not
only defined for differential operators but also for recurrence operators, i.e.,
elements of an Ore algebra of the form $R[x]\<Sx>$ or~$K(x)\<Sx>$. Some other
methods defined for recurrence operators are listed in the following table.

\begin{center}
  \begin{tabular}{|l|p{.55\hsize}|}
    \hline
    Method name & short description \\\hline
    \verb|to_D()| & converts annihilator for the coefficients in a power series to a
        differential operator for the sum \\
    \verb|to_F()| & converts shift operator to a difference operator\\
    \verb|to_T()| & converts to a differential operator in terms of the Euler derivative \\
    \verb|annihilator_of_sum()| & converts an annihilator for $f(n)$ to one for the sum $\sum_{k=0}^n f(k)$\\
    \verb|annihilator_of_composition()| & converts an annihilator for $f(n)$ to one for $f(\lfloor un+v \rfloor)$ where $u, v \in \set Q$ \\
    \verb|annihilator_of_interlacing()| & interlaces two or more sequences\\
    \verb|desingularize()| & computes a left multiple of this operator with polynomial coefficients and 
           lowest possible leading coefficient degree\\
    \verb|associate_solutions(p)| & applied to an operator~$P$, this computes, if possible, an operator $M$
    and a rational function $m$ such that $(S-1)M=p+mP$ (see \cite{abramov99} for further information)\\
    \verb|polynomial_solutions()| & computes the polynomial solutions of this operator\\
    \verb|rational_solutions()| & computes the rational function solutions of this operator\\
    \verb|generalized_series_solutions()| & computes asymptotic expansions of sequences annihilated by the operator\\ 
    \verb|to_list()| & computes terms of a sequence annihilated by the operator\\\hline
  \end{tabular}
\end{center}

@@

As an example application, we compute an annihilator for the sequence
$c(n) = \sum_{k=0}^n 1 / k!$:

\begin{sageexample}
  sage: R.<n> = ZZ[]; A.<Sn> = OreAlgebra(R)
  sage: inverse_factorials = (n + 1) * Sn - 1
  sage: partial_sums = inverse_factorials.annihilator_of_sum()
  sage: partial_sums
\end{sageexample}

@@

\subsection{Guessing}

Guessing is, in some sense, the reverse operation of \verb|to_list| for
recurrence operators, or \verb|power_series_solutions| for differential
operators.  It is of the most popular features of packages like gfun, and there
are even some special purpose packages dedicated to this
technique~\cite{kauers09a,hebisch11}. The basic idea is simple. If we are given
a finite array of numbers, thought of as the first terms of an infinite
sequence, we may want to know whether this sequence satisfies a recurrence. The
algorithm behind a guessing engine searches for small equations matching the
given data.  Generically, no such equations exist, so if some are found, it is
fair to ``guess'' that they are in fact valid equations for the whole infinite
sequence.

@@

The method tests a sequence of points $(r_1,d_1)$, $(r_2,d_2)$, \dots\ right
below this second hyperbola. Success at a point $(r_i,d_i)$ means that some
evidence for an operator of order $\leq r_i$ and degree $\leq d_i$ has been found. 
This operator however is not explicitly computed. Instead, the method uses the
partial information found about this operator to calculate an operator which with 
high probability is the minimal order operator satisfied by the sequence in question. 
This operator is usually more interesting than the one at $(r_i,d_i)$, and its 
computation is usually more efficient.

@@
